---
title: "Introduction to Statistics for Biosciencies"
output: html_notebook
---

###### Course **CIIMAR**
###### Lecturer **Aldo Felpeto**
###### 11-14 Jan 2021   
### **11 Jan - Session 1**

# DESCRIPTIVE STATISTICS   

```{r}
x <- c(3,6,7,4,3,2,5,4,3,3,7,8,3)
```

```{r}
library("modeest")
mlv(x, method = "mfv")
```

```{r}
hist(x)
```

```{r}
x2 <- rnorm(50)
x2
```

```{r}
plot(x2)
```
```{r}
hist(x2)
```

### MEASURES OF CENTRAL TENDENCY

```{r}
mlv(x2,method="naive",bw=1/3) # finds interval with more observations (and considers the middle point as the mode)
```
```{r}
mlv(x2,method="parzen",kernel="gaussian")
```

#### Median
Not sensitive to outliers
```{r}
x3<-c(3,6,NA,7,4,3,2,5,4,3,NA,3,7,8,3)
median(x3,na.rm=TRUE) # NOT sensitive to outliers
```


#### Arithmetic Mean
```{r}
mean(x3, na.rm=TRUE) # sensitive to outliers
```

```{r}
mean(x2,na.rm=TRUE,trimm=0.1) # truncated mean (trimming data) remove outliers- here removing 0.1 or 10% from each tail of the distribution
```

### Means for vars that by nature are not additive or linear, but multiplicative, exponential or logarithmic

#### Geometric Mean   
Geometric mean good for rates of change
```{r}
library(psych)
geometric.mean(x3)
```

#### Harmonic Mean   
For ranges with high magnitudes (not for rates of change) 
```{r}
library(psych)
harmonic.mean(x3)
```

### MEASURES OF DISPERSION   
```{r}
z <- rnorm(13)
z
```

#### Variance    
Variance (squared values of vars) (since square no negative values) (so used in models)
```{r}
var(z) # 
```

#### Standard Deviation   
Square root of variance (dispersion in same scale of vars) (better than var() to describe the data so easier to interpret)
```{r}
sd(z)
```
sd for data n < 13 sd should be calculated dividing by n-1 bather than dividing by n.


*Never use **Standard Error** *to describe data. It is useful for estimation of parameters. So for inferential, not descriptive stats.*     


#### Coefficient of Variation   
Useful to compare vars which have different scales.
Divide standard deviation by the mean: cv = sd/mean
```{r}
cv <- sd(z)/mean(z)
cv
```

#### Interquartile Range
```{r}
a <- rnorm(100)
a
```

```{r}
IQR(a)
```

### MEASURES OF ASSOCIATION   
Measure the variance of the level of association between variables.   

#### Correlation Coefficient   
**Parametric Vs Non-Parametric**   

### Parametric   
Assumes that the data follows certain parameters:   
 - data follows a specific type of distribution (Gaussian, Poisson, Gamma, etc.)   

#### Perason's Correlation Coefficient (Parametric)   
Assumptions:   
- data is normally distributed;   
- there is a linear dependence.   



### Non-Parametric    
No assumptions of normality or linearity.   
One assumption:   
 - Data is **monotonic** (always increase or decrease)   
Used for ranked data.   

#### Spearman's
For ranked data. Can also be used for ordinal variables but only with large data sets.  


#### Kendal   
For ordinal variables with small data sets.      


##### Compare Coefficients   
```{r}
x<-rnorm(30)
x
```
```{r}
y1 <- (0.7*x+0.3)+rnorm(30,0,0.25) # rnorm() to add noise
y1
```

```{r}
plot(x,y1)
```
###### Sigmoid function   
```{r}
y2<-(1/(1+exp(-2.5*x)))+rnorm(30,0,0.05)
y2
```

```{r}
plot(x, y2)
```

#### Good data for pearson   
```{r}
cor.test(x,y1,method="pearson") 
```
```{r}
cor.test(x,y1,method="spearman")
```
```{r}
cor.test(x,y1,method="kendall")
```

#### Good data for Spearman's
```{r}
cor.test(x,y2,method="pearson") # good data for spearman
```
```{r}
cor.test(x,y2,method="spearman")
```
```{r}
cor.test(x,y2,method="kendall")
```




# PROBABILITY DESTTRIBUTIONS    
The probability of a possible outcome (value)

### Gaussian (Normal)   
dnrom(x)

### Poisson distribution   
Positive integers (discreet, so not continuous)   
dpois(x)    

### Gamma Distribution   
Always positive and continuous   


### Chi Square   
`r dchisq(18.3,10)` 
Functions that start with **d** (density): Statistical probability at specific point (18.3 in this case, 10 is degrees of freedom). This is the p-value

`r pchisq(18.3,10)` 
Functions that start with **p** (probability): Integral of probability, or sum up to that point

```{r}
library(sjPlot)
dist_chisq(deg.f=10)
```

# DATA EXPLORATION   
```{r}
summary(x2)
```


#### Helping functions for checking for completeness of data, namely missing values   
```{r}
library(pastecs)
stat.desc(x2) 
```

```{r}
require(psych)
describe(x2)
```

```{r}
head(airquality)
```
```{r}
summary(airquality)
```

#### Correlation between several vars
```{r}
pairs(na.omit(airquality[1:4]), panel = function(x, y) {
    points(x, y)
    lines(lowess(x, y), lwd = 2, col = "red")
    })
```

#### histograms
```{r}
par(mfrow=c(2,2))
hist(airquality$Ozone)
hist(airquality$Solar.R)
hist(airquality$Wind)
hist(airquality$Temp)
```

*log does not make vars more normal in general, makes SOME. Other could do with dif. transformation*   

```{r}
par(mfrow=c(2,2))
hist(log(airquality$Ozone))
hist(log(airquality$Solar.R))
hist(log(airquality$Wind))
hist(log(airquality$Temp))
```


#### Summary per var (months)      
```{r}
tapply(airquality$Ozone,airquality$Month,summary)
```

```{r}
par(mfrow=c(2,2))
boxplot(airquality$Ozone~airquality$Month)
boxplot(airquality$Solar.R~airquality$Month)
boxplot(airquality$Wind~airquality$Month)
boxplot(airquality$Temp~airquality$Month)
```

```{r}
library(ggplot2)
airquality$Month<-as.factor(airquality$Month)
ggplot(airquality,aes(x=Month,y=Ozone,fill=Month)) +
    geom_violin()
```


# EXERCISE 1
In the file Cholesterol.csv, we have the measurements of Cholesterol in blood in a number of patients. It was recorded as well the age, the gender and the physical
condition (attributed to 4 different categories) of the patients. Calculate the descriptive statistics that you consider appropriate for these data and perform basic
plots to have a look at the relationships and distributions of the variables.

```{r}
data <- read.csv2("Cholesterol.csv")
str(data)
```
```{r}
summary(data)
```

### Cholesterol
```{r}
mean(data$Chol.Blood)
```
```{r}
median(data$Chol.Blood)
```
```{r}
sd(data$Chol.Blood)
```



### Gender   
```{r}
plot(as.factor(data$Gender), data$Chol.Blood)
```

```{r}
female <- data[data$Gender == "F", ]
male <- data[data$Gender == "M", ]
```
```{r}
mean_female <- mean(female$Chol.Blood)
mean_female
```
```{r}
mean_male <- mean(male$Chol.Blood)
mean_male
```
### Hospital
```{r}
plot(as.factor(data$Hospital), data$Chol.Blood)
```

```{r}
library(dplyr)
by_hosp <- data %>%
            group_by(Hospital) %>%
            summarise(n = n(),
                      mean_chol = mean(Chol.Blood),
                      sd_chol = sd(Chol.Blood),
                      mean_age = mean(Age),
                      sd_age = sd(Age))

by_hosp
```

### Physical Condition
```{r}
plot(as.factor(data$Physical.condition), data$Chol.Blood)
```

### Gender
```{r}
rl <- lm(data = data, Chol.Blood ~ Age)
plot(data$Age, data$Chol.Blood)
# abline(rl[[1]][1], rl[[1]][2], col = 'red')
abline(rl, col = 'red')
```

##### Check for Outliers on the Model
```{r}
library(car)
outlierTest(rl)
```























[link](http://rmarkdown.rstudio.com)
